---
title: "Thesis Preliminary Analysis"
author: "CRUSTOM JAKE D. RONCAL"
format: 
  html:
    self-contained: true
    mainfont: arial
    fontsize: 1em
    linestretch: 1.5
    toc: true
    toc-location: left
    number-sections: true
    code-fold: true
    theme:
      light: cosmo
      dark: darkly
execute:
  message: false
  warning: false
  fig-align: center
editor: visual
---

# Panel Data Analysis on the Effects of Economic Growth and Policy on Deforestation Rates in Selected ASEAN Member States

## Brief Rationale

Forests are crucial for maintaining ecological harmony and rich biodiversity. They supply a myriad of essential benefits, including the production of oxygen, storage of carbon, and a place to live for countless variety of species, which accounts for roughly 80% of the world’s terrestrial biodiversity. Moreover, forests play an important role in the global carbon cycle as they act as filters of natural waters, climate reduction by absorption of carbon dioxide, and soil conservation (Food and Agriculture Organization, 2010). From an economic perspective, forests also have a significant role to play by providing natural endowments to the lives of millions. A report in 2010 revealed that approximately 1.6 billion of the world’s population relies heavily on forests for their source of living, survival, and holistic welfare (FAO, 2010). Forests are one of the most precious resources that Asia possesses, as they consist of 17% of the world’s forest cover. However, due to the abundance of resources in the region, Asia is currently under an alarming threat of deforestation, driven specifically by illegal logging, expansion of used agricultural land, and infrastructure development.

## Objectives

This study will empirically analyze how economic and institutional determinants will affect the rate of deforestation across selected ASEAN member states by incorporating the Environmental Kuznets Curve, primarily focusing on the relationship between economic growth and deforestation using panel data regression. Specifically,

1.  to describe the trends of the rate of deforestation and GDP per capita across selected ASEAN member states from 2008-2020

2.  to assess the significance and magnitude of economic growth and policy factors influencing deforestation rates in selected ASEAN members states; and

3.  to add sustainable policy insights regarding the effects of extensive economic activities and quality of institutional policies on the state of forest among ASEAN countries

## Data Analysis

**Dependent Variable:**

-   Rate of Deforestation

**Independent Variable:**

-   GDP per Capita

-    Foreign Direct Investment

-   Wood Exports

-   Share of Renewable Energy ; and

-   Control of Corruption

### Descriptive Statistics

```{r}
#Read the dataset
thesis <- read.csv("~/data.csv")

# omit missing values
thesis <- na.omit(thesis)

# Load necessary libraries
library(stargazer)

#Drop Variables
thesis <- thesis[, !(names(thesis) %in% c("GDPpc2","PopD","Trade", "AgriLU", "RegQ", "RLaw"))]

# Create a clean Descriptive Statistics Table
stargazer(
  thesis[, !(names(thesis) %in% c("Country", "Year", "GDPpc2"))],  # Remove unwanted variables
  type = "text",        
  title = "Descriptive Statistics of Key Variables", 
  digits = 3,           # Number of decimals
  summary.stat = c("mean", "sd", "min", "median", "max"),  # Statistics you want
  out.header = FALSE,   # No extra header
  initial.zero = TRUE   # Show leading 0 before decimal
)
```

Table \_ shows the descriptives statistics of the variables used in the study, which utilized five ASEAN Member States—Indonesia, Malaysia, Philippines, Thailand, and Vietnam—spanning the from the period 2008-2020. It shows the number of observation (N), mean, standard deviation (St. Dev), minimum values, and maximum values from each variables examined the study. Moreover, the variables used in the study were rate of deforestation (RDef), GDP per capita (GDPpc), Share of Renewable Energy (RENEW), Wood Exports (WoodEx), Foreign Direct Investment (FDI), and Control of Corruption (CCor).

**Rate of Deforestation.** The rate of deforestation (RDef) which is the main dependent variable of the study, measured as the average annual change, has a mean value of -0.119 and standard deviation of 0.685. The negative mean suggests, that on average, the country covered in this study are experiencing moderate reduction in deforestation over period of the study, which may also suggest a slow transition to reforestation. However, the substantial difference between the minimum value of -2.246 and maximum value of 1.388 implies a significant variation across ASEAN countries over year. Some countries may have strengthened their national efforts to restore the forest areas (negative values) whereas other countries are may still extracting natural resources to meet their national needs.

**GDP per Capita.** Gross Domestic Product per Capita (GDPpc), measured in current USD, presents a mean value of 12,712.940 with a standard deviation of 6,748.938. This shows how dispersed the income per capita is across selected ASEAN member states, which may also reflect the variation in the level of economic development. The minimum GDP per capita is listed at 4,916.131, while maximum value peaked at 29,750.270. The noticeable range in per capita income may suggest that countries included the study are in both lower-middle income (Vietnam, Philippines,) and upper-middle income economies (Malaysia, Thailand). This spread in income may have a significant effect in regulating environmental effects.

**Share of Renewable Energy Consumption.** Renewable energy consumption (RENEW), expressed as the percentage of total final energy consumption, presented an average of 23.669% across observations, with a relatively high standard deviation of 11.374%. The minimum value was 2.000% while maximum value settled at 41.100%. This highly dispersed values from the observation suggest that while there are ASEAN member states that already relied heavily on renewable energy, there are still countries that settled on consuming non-renewable energy.  Share of renewable energy consumption patterns could have significant impact on deforestation rates since transitioning to a cleaner source of energy consumption may mitigate forest and natural resources degradation caused by intensive traditional biomass utilization.

**Wood Exports.** Wood exports (WoodEx), which represents the share of wood products in merchandise exports, tallied an average on 1.998% across sample on a standard deviation of 0.313. The minimum value is recorded at 1.420% whereas maximum value is set at 2.530%, suggesting that while it is an important sector, it recorded a relatively lower variation and fluctuation across time. However, it is interesting to identify how exporting wood products, especially in countries reliant on forest resources like developing countries, impact the degradation of natural resources

**Foreign Direct Investment.** Foreign direct investment (FDI) inflows, expressed as a percentage of GDP, shows a mean value of 5.076% and a standard deviation of 4.681. FDI in selected ASEAN states reached the highest value at 0.21675% and lowest at 0.724%. The widespread in FDI inflows may signify variation in economic openness among ASEAN states, potentially affecting the regulation of land-use practices, like extraction of natural resources either through direct investment in resources-depleting industries or other investment options related to environmental utilization.

**Control of Corruption.** Control of Corruption (CCor), showed a mean value of -0.399 corresponding to a standard deviation of 0.306, ranging from a minimum value of -0.890 to a maximum value of 0.397.  This variable is measured on an idex where higher values indicate better of control of corruption practices while lower values show poor regulation of corruption practices. The negative average may be inferred as countries involved in the study are experiencing difficulty containing corruption. Corruption, when not controlled effectively, may have significant effect on the degradation of forest due to poor land administration, lack of policies on logging practices, and enforcement of sustainable environmental practices.

### Converting the Data in Panel Dataset

```{r}
#converting the data into a panel dataset

library(plm)
library(stargazer)

paneldata <- pdata.frame(thesis, index = c("Country", "Year")) 
pdim(paneldata)

stargazer(paneldata, type = "text", title = "Descriptive Statistics of Panel Data", digits = 3) 
```

### Exploratory Descriptive Analysis

```{r}
#exploratory descriptive analysis
library(dplyr)
library(ggplot2)

#logging GDPpc and RDef
log <- thesis %>% mutate(log_RDef = RDef, log_GDPpc = log(GDPpc))
```

#### A. Trend Analysis of the Variables from 2008-2020

```{r}
#Rate of Deforestation
ggplot(log, aes(x = Year, y = log_RDef, color = Country)) +
  geom_line() +
  facet_wrap(~ Country, scales = "free_y") +
  scale_x_continuous(breaks = c(2010, 2015, 2020)) +
  theme_minimal() +
  ggtitle("Rate of Deforestation Over Time by Country") +
  xlab("Year") +
  ylab("Rate of Deforestation")
```

**Rate of Deforestation** *(%)*. Indonesia started with a lower deforestation rate but experienced an immediate surge in 2011 while maintaining a stable pattern up until 2015 before it declined heavily the following year, peaked again in 2017 before meeting a lower stable pattern in the late 2017 onwards. The fluctuations shows that Indonesia had inconsistencies containing the decline of forests and regulating policies to combat the extractions. Malaysia has experienced a sharp decline of deforestation rate starting from year 2011, maintaining a stable pattern until 2015 before the nation experienced a sharp increase in 2016 and controlling to a lower rate the following years. The stable pattern from 2011-2015 indicates that the nation has shown a temporary yet potential mitigation of forest extraction or better land-use practices. Philippines has evidently showed a steep drop of deforestation rates in 2010 and maintained a lower deforestation rate since then, indicating that the Philippines has been practicing environmental conservation and forest-related policies. Thailand presents a moderately increasing pattern of deforestation rates from 2008-2020, potentially indicating that while the rate is still below 1%, the nation should regulate policies to control the slowly increasing extraction of forest products in Thailand. Lastly, Vietnam is experiencing a negatively fluctuating trends of deforestation rates that even hit up to -2% in 2015, which may indicate that the nation has long been in the stages of reforesting the extracted lands and controlling the forests removal.

```{r}
#GDP per capita
ggplot(log, aes(x = Year, y = log_GDPpc, color = Country)) +
  geom_line() +
  facet_wrap(~ Country, scales = "free_y") +
  scale_x_continuous(breaks = c(2010, 2015, 2020)) +
  theme_minimal() +
  ggtitle("GDP per capita over time by Country") +
  xlab("Year") +
  ylab("GDP per capita")
```

**GDP per Capita** *(thousand USD)*. As evident in the graph, all ASEAN nations included in the study showed an upward trend in GDP per capita from 2008-2020 though possessed different economic wealth ranges. Indonesia has seen to have a steady increase in GDP per capita from 8.8 USD in 2008 to a peaking 9.4 USD in 2019. Malaysia also presented a peaking trend but at a moderately higher level starting at 9.8 USD in 2008 to a consistent increase of 10.3 USD in 2019. Philippines exhibited a stable economic progress with an 8.6 USD GDP per capita in 2008 to a slightly higher 9.0 value in 2019. Thailand also showed a pronounced economic growth ranging from 9.4 USD in 2008 to a 9.9 USD GDP per capita in 2019. Finally, Vietnam has also experienced similar upward trend but in a lower range, starting at 8.50 USD to 9.25 USD in 2019. Noticeably, all countries except Vietnam experienced a downturn in GDP per capita in 2020 potentially because of the occurrence of the global pandemic that shut down international economic activities.

```{r}
#Share of Renewable Energy
ggplot(log, aes(x = Year, y = RENEW, color = Country)) +
  geom_line() +
  facet_wrap(~ Country, scales = "free_y") +
  scale_x_continuous(breaks = c(2010, 2015, 2020)) +
  theme_minimal() +
  ggtitle("Share of Renewable Energy Over Time by Country") +
  xlab("Year") +
  ylab("Share of Renewable Energy")
```

**Share of Renewable Energy** *(% of TFEC)***.** Selected ASEAN member states demonstrated varying trends of share of renewable energy over time. Indonesia has observed a significant decline in the renewable energy consumption starting at 40% in 2008 and dropping to about 20% in 2020. This indicate that the nation has been once highly reliant on renewable energy resources but then slowly shifted to the usage of nonrenewable energy resources such coal and natural gas. Conversely, Malaysia exhibited a low yet gradual dependence in renewable energy resources over time ranging at about 2% in 2008 to a steadily increasing 7% in 2020. Philippines, while presented a declining pattern, maintained the highest renewable energy share among selected ASEAN member states striking at 34% in 2008, declined momentarily at 32% in 2010, increased again at 34% in 2011 before experiencing a consistent declining trend the following years until 2019 with a 27% and seen a slight upward in 2020. Thailand showed fluctuating patterns over time recording an almost 23% renewable energy share in 2008, peaking at more than 24% in 2015, declined severely at 22% in 2016, increased again in the latter part in 2019 but then dropped again the year after. The fluctuation may be due varying shifts of regulatory incentives given to renewable energy consumption. Lastly, Vietnam exhibited a declining pattern just like that of Indonesia and Philippines. The nation recorded an almost 40% reliance on renewable energy consumption in 2008 dropping at 20% in 2020. Similarly, this can be attributed to the nation’s high reliance on nonrenewable energy resources like fossil fuels.

```{r}
#Foreign Direct Investment
ggplot(log, aes(x = Year, y = FDI, color = Country)) +
  geom_line() +
  facet_wrap(~ Country, scales = "free_y") +
  scale_x_continuous(breaks = c(2010, 2015, 2020)) +
  theme_minimal() +
  ggtitle("Foreign Direct Investment Over Time by Country") +
  xlab("Year") +
  ylab("FDI")
```

**Foreign Direct Investment** *(% of GDP)*. All ASEAN member states involved in the study observed fluctuating FDI pattern over the period of the analysis. Indonesia recorded the highest percentage of 4% in 2015 before experiencing a steady decline and settling at almost 2% in 2020.  The trend showed that Indonesia has not been dependent on foreign investments for local industrialization. Identically, Malaysia showed a fluctuating trend over time peaking at 3% in 2011 and 2015 but then declined in 2020. Philippines showed the highest reliance on FDI among ASEAN member states showing an upward trajectory with a 5% value in 2010, peaking at more than 20% in 2015 with a moderate decline in between years, and decreasing slightly in 2020, This indicates a high foreign attractiveness on the nation’s economic affairs and enhanced regulation on trade and investments. Thailand has experienced a sharp peak in late 2000’s at 7.5% before declining to less than 2.5% the following year, peaked again at 5% in 2015 and declining again at 2.5% in 2020. Vietnam peaked at 8% in 2010 but experienced sharp decline at 2.5% in 2015 and eventually experienced fluctuations up until 2020. The fluctuations show that nations are experiencing dynamic foreign investment attractiveness due to possible political and economic restrictions.

```{r}
#Wood Exports
ggplot(log, aes(x = Year, y = WoodEx, color = Country)) +
geom_line() +
  facet_wrap(~ Country, scales = "free_y") +
  scale_x_continuous(breaks = c(2010, 2015, 2020)) +
  theme_minimal() +
  ggtitle("Wood Exports Over Time by Country") +
  xlab("Year") +
  ylab("Wood Exports")
```

**Wood Exports** *(thousand USD).* Fluctuations have also been evident on the wood exports of the nations covered. Indonesia experienced high wood exports in 2010 at almost 2.4 USD but then followed sharp decline of less than 2.0 USD in 2011 before experiencing a peaking wood export in 2016 onwards with fluctuations in between. This suggests that there is a slow yet unclear demand for woods from the nation over time. Indonesia registered more than 1.7 USD in 2008 declined steadily to less than 1.5 USD in 2015 increased again at almost 1.8 USD in 2017 and recorded increasing exports since then, which may suggest that Malaysia is slowly exporting larger portions of wood products over time. Philippines has seen to have an increasing wood exports pattern over time ranging from 2.0 USD in 2008 to more than 2.4 USD in 2020 indicating an increasing exportation of wood products to international markets. Thailand also experiences inconsistent patterns over time with high wood exports on years 2010 with more than 1.8 USD but decline in the succeeding year before peaking again in 2015 with more than 1.7 USD and declining back to less than 1.6 USD in late 2019. The fluctuations may be due to shifts in forestry and trading agreements in between years. Vietnam has recorded a peak point at 2010 with more than 2.4 USD worth of wood exports followed before declining wood exports the following years and registering less than 2.0 USD in 2015 but then showed gradual increase to 2.2 USD in 2020. This suggests a declining exports of wood products from 2010 to 2015 due to enhanced local trading policies.

```{r}
#Control of Corruption
ggplot(log, aes(x = Year, y = CCor, color = Country)) +
  geom_line() +
  facet_wrap(~ Country, scales = "free_y") +
  scale_x_continuous(breaks = c(2010, 2015, 2020)) +
  theme_minimal() +
  ggtitle("Control of Corruption Over Time by Country") +
  xlab("Year") +
  ylab("Control of Corruption")
```

#### B. DV-IV Visual Presentation

```{r}
library(ggplot2)
library(dplyr)
library(scales)
```

```{r}
#RDef-GDPpc
ggplot(log, aes(x = log_GDPpc, y = log_RDef)) +
  geom_line(aes(group = Country), color = "gray60", linewidth = 0.8) +  # Connect points
  geom_point(color = "gray30", size = 2) +
  facet_wrap(~ Country, scales = "free") +
  labs(
    title = "Relationship between Deforestation Rates and GDP per Capita",
    subtitle = "Per Country (ASEAN Panel Data)",
    x = "GDP per Capita",
    y = "Deforestation Rate"
  ) +
  theme_classic(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "grey90", color = "grey50"),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    panel.spacing = unit(1, "lines")
  )
```

```{r}
#RDef-FDI
ggplot(log, aes(x = FDI, y = log_RDef)) +
  geom_line(aes(group = Country), color = "gray60", linewidth = 0.8) +  # Connect points
  geom_point(color = "gray30", size = 2) +
  facet_wrap(~ Country, scales = "free") +
  labs(
    title = "Relationship between Deforestation Rates and FDI (% of GDP)",
    subtitle = "Per Country (ASEAN Panel Data)",
    x = "Foreign Direct Investment (% of GDP)",
    y = "Deforestation Rates"
  ) +
  theme_classic(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "grey90", color = "grey50"),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    panel.spacing = unit(1, "lines")
  )
```

```{r}
#RDef-RENEW
ggplot(log, aes(x = RENEW, y = log_RDef)) +
  geom_line(aes(group = Country), color = "gray60", linewidth = 0.8) +  # Connect points
  geom_point(color = "gray30", size = 2) +
  facet_wrap(~ Country, scales = "free") +
  labs(
    title = "Relationship between Deforestation Rates and Share of Renewable Energy",
    subtitle = "Per Country (ASEAN Panel Data)",
    x = "Share of Renewable Energy Consumption",
    y = "Deforestation Rates"
  ) +
  theme_classic(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "grey90", color = "grey50"),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    panel.spacing = unit(1, "lines"))
```

```{r}
#RDEf-WoodEx
ggplot(log, aes(x = WoodEx, y = log_RDef)) +
  geom_line(aes(group = Country), color = "gray60", linewidth = 0.8) +  # Connect points
  geom_point(color = "gray30", size = 2) +
  facet_wrap(~ Country, scales = "free") +
  labs(
    title = "Relationship between Deforestation Rates and Wood Exports",
    subtitle = "Per Country (ASEAN Panel Data)",
    x = "Wood Exports",
    y = "Deforestation Rates"
  ) +
  theme_classic(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "grey90", color = "grey50"),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    panel.spacing = unit(1, "lines")
  )
```

```{r}
#RDef-CCor
ggplot(log, aes(x = CCor, y = log_RDef)) +
  geom_line(aes(group = Country), color = "gray60", linewidth = 0.8) +  # Connect points
  geom_point(color = "gray30", size = 2) +
  facet_wrap(~ Country, scales = "free") +
  labs(
    title = "Relationship between Deforestation Rates and Control of Corruption",
    subtitle = "Per Country (ASEAN Panel Data)",
    x = "Control of Corruption",
    y = "Deforestation Rates"
  ) +
  theme_classic(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "grey90", color = "grey50"),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    panel.spacing = unit(1, "lines")
  )
```

### Regression Analysis

#### A. Pooled OLS

```{r}
#A. Pooled OLS
Pooled <- plm(RDef ~ GDPpc + FDI + WoodEx + RENEW + CCor, data = paneldata, index = c("Country", "Year"), 
                model = "pooling")
summary(Pooled)
```

#### B. Random Effects (Wallace-Hussain Transformation)

```{r}
#B. Random Effects (Wallace-Hussain Transformation)
Random <- plm(RDef ~ GDPpc + FDI + WoodEx + RENEW + CCor, data = paneldata, index = c("Country", "Year"), 
              model = "random", random.method = "walhus")
summary(Random)
```

#### C. Fixed Effects

```{r}
#C. Fixed Effects
Fixed <- plm(RDef ~ GDPpc + FDI + CCor + WoodEx + RENEW , data = paneldata, index = c("Country", "Year"), 
             model = "within") 
summary(Fixed)
```

#### D. Regression Tables

```{r}
#Generate Well-Formatted Table with Proper Labels
stargazer(Pooled, Random, Fixed, type = "text",
          title = "Panel Regression Results",
          column.labels = c("Pooled", "Random", "Fixed"),
          digits = 3, align = TRUE)
```

#### E. Choosing the Best Model to Use

```{r}
# 1. Breusch-Pagan Lagrange Multiplier test (Testing: Pooled OLS vs. Random Effects)
plmtest(Random, type = "bp")
```

Null Hypothesis (H₀): There are no panel (individual-specific) effects. Pooled OLS is appropriate.

Alternative Hypothesis (H₁): There are significant panel effects. Random Effects model is preferred.

```{r}
# 2. Pooled vs Fixed Effects (Testing: Pooled OLS vs. Fixed Effects)
pFtest(Fixed, Pooled)

```

Null Hypothesis (H₀):All individual effects are zero (no fixed effects needed). Pooled OLS is appropriate.

Alternative Hypothesis (H₁): At least one individual effect is non-zero (Fixed Effects model is needed).

```{r}
# 3. Fixed vs Random Effects (Hausman Test)
phtest(Fixed, Random)
```

Null Hypothesis (H₀):Random Effects estimator is consistent and efficient → RE is appropriate.

Alternative Hypothesis (H₁):Random Effects estimator is inconsistent → prefer Fixed Effects.

### Post Estimation Analysis

```{r}
library(lmtest)
library(plm)
library(car)
```

#### Variance Inflation Factor

```{r}
#A. Variance Inflation Factor
vif(lm(RDef ~ GDPpc + RENEW + WoodEx + FDI + CCor, data = paneldata))
```

Purpose: Detect multicollinearity among the explanatory variables.

Rule of Thumb: VIF \> 10: High multicollinearity (problematic)

#### Breusch-Pagan Test for heteroscedasticity

```{r}
bptest(Random)
```

Purpose: Test if residuals have constant variance (homoskedasticity).

Null Hypothesis (H₀): Homoskedasticity (constant variance of errors) Alternative Hypothesis (H₁): Heteroskedasticity (variance of errors is not constant)

#### Breusch-Godfrey/Wooldridge test for serial correlation in panel models

```{r}
pbgtest(Random)
```

Purpose: Checks autocorrelation in the panel data errors. Null Hypothesis (H₀): No serial correlation in the idiosyncratic errors Alternative Hypothesis (H₁): Serial correlation exists

#### Cross-sectional Dependence (Pesaran's CD test)

```{r}
pcdtest(Random, test = "cd")
```

Purpose: Check if errors across cross-sectional units (countries) are correlated.

Null Hypothesis (H₀): No cross-sectional dependence (errors are independent across panels)

Alternative Hypothesis (H₁): Presence of cross-sectional dependence
